Cluster Dynamics Using the Hobby Eberly Telescope Dark Energy Experiment

### 2015-06-10 ms.tex
updated
“finished” introduction, reorganized a bit, added section about
membership, and much more.

\section{INTRODUCTION}
The Hobby Eberly Telescope Dark Energy eXperiment (HETDEX; \citealt{Hill2008}) is a trailblazing effort to observe high-redshift large scale structures using cutting edge wide-field integral field unit (IFU) spectrographs. Designed to probe the evolution of the dark energy equation of state etched onto high redshift ($z>2$) galaxies by the Baryon Acoustic Oscillations \citep{Eisenstein2005} in the first moments of the universe, the survey will observe two fields for a total of 420 \degsq\ from two fields (300 \degsq, Spring field and 120 \degsq, Fall field). Tuned to find \lya\ emitting (LAE) galaxies at $1.9<z<3.5$, HETDEX expects to find 800,000 LAEs, and more than one million [\ion{O}{ii}] emitting galaxies at $z<0.5$ masquerading as high-redshift galaxies.

While a large portion of thee $\sim10^6$ interloping [\ion{O}{ii}] galaxies will be field (not associated with a larger structure) galaxies, the large area covered by HETDEX is expected to contain as many as 100 Virgo-sized ($M_{dyn}\sim 10^{15}$ \msol) clusters at $z<0.5$ (\editorial{citation?}). The near-complete spectroscopic coverage allows an unprecedentedly detailed look at a large number of clusters ranging from group scales to the very massive. In addition to the recovery of accurate dynamical masses, detailed investigations of the of dynamical state of the clusters is possible.

In addition to the spectroscopy, an imaging campaign is require first identify galaxy over densities. HETDEX overlaps with the Sloan Digital Sky Survey (SDSS; \citealt{Blanton2001a}), SDSS strip 82 \citep{Annis2014}, the Dark Energy Survey \citep{DES2005}, and the upcoming DECam/IRAC Galaxy Environment Survey (DIRGES; PI: Papovich, C. Papovich \etal in preparation).

This support will allow Mr. Boada to contribute to the core science of my NSF grant (AST 1413317) for the DECam/IRAC Galaxy Environment Survey (DIRGES). He will studying both simulated and real data to understand the reliability in measuring galaxy cluster masses and dynamical states using spectroscopic information for emission-line galaxies. DIRGES will identify as many as 100 galaxy cluster progenitors at 0.8 $<$ z $<$ 1.5. Significant gains in the understanding of these clusters are possible if we can (1) interpret their dynamical nature through spectroscopy, and (2) if we can link them to clusters at lower redshift, z $<$ 0.5, from within the same dataset to suppress potential observational biases. I have plans to obtain the spectroscopy to achieve both of these supplemental goals, but the majority of the spectroscopic redshifts will be for emission-line galaxies. It is not straightforward to use these data to interpret the kinematic and dynamical states of the clusters. DIRGES is covered by SDSS Stripe 82 \citep{Annis2014}, the Dark Energy Survey (DES), and the Hobby Eberly Telescope Dark Energy eXperiment (HETDEX). One key is HETDEX, which is a spectroscopic survey covering $\sim$400 deg$^2$ and measuring redshifts for as many as ∼106 emission-line galaxies at z $<$ 0.5. HETDEX will contain as many as 100 galaxy Virgo-sized clusters ($M_{dyn}\sim10^{15}\msol$) out to z $<$ 0.5. And, because HETDEX covers DIRGES, the data will contain both low-redshift and high-redshift clusters in the same imaging dataset, allowing the most homogeneous study of the low-z and high-z universe yet.

\editorial{The text below is taken from Chiang 2015 and is NOT original to me.}

In a 3-year baseline starting from late 2015, HETDEX will leverage the cosmic evolution of the dark energy equation of state with high redshift (z $>$ 2) constraints imprinted by the Baryon Acoustic Oscillations (BAO; Eisenstein 2005) in the early universe. The program will perform a redshift survey of LAEs in 300 deg2 (Spring field) plus 150 deg2 (Fall field) at 1.9 $<$ z $<$ 3.5 (with a filling factor of 1/4.5), with a total survey volume of $\sim$ 8.5 Gpc3. The survey uses the 10-m Hobby-Eberly Telescope (HET; \citealt{Ramsey1998}) with a wide-field upgrade to reach a 22x22 arcmin2 field-of-view. Blind spectroscopy (R $\sim$ 750 in $350–550$ nm) with no pre-selection of targets will be performed using the Visible Integral-Field Replicable Unit Spectrograph (VIRUS; \citealt{Hill2012}). With the LAE redshifts, HETDEX will pinpoint numerous locations of the highest density concentrations at 1.9 $<$ z $<$ 3.5, generating a substantially large and homogeneous sample of cluster progenitors in the key epoch of cluster formation before virialization.

@@

The ``Aardvark'' mock galaxy catalogs (R. Wechsler et al., private communication) are derived from a semi-analytical model (SAM) tied to the Bolshoi cosmological simulation \citep{Klypin2011} and designed for DES. They provide a 10313 sq. degree (one quarter of the sky) catalog to full DES depth and contains 1.36 billion galaxies which have a signal-to-noise of at least five in one or more DES observing bands. The catalog footprint is shown in Figure~\ref{fig:survey area}.

@@

\subsection{Mock Observations}
\begin{figure} 
	\includegraphics[width=0.49\textwidth]{f01.pdf} 
	\caption{Representative observation tiling scheme for the HETDEX $16'$ pointings. Each colored square is a single VIRUS IFU and the dashed octagons approximate the size of a single observation. See the text for more details.} \label{fig:ifu layout} 
\end{figure}

Our ``observations'' consist of placing masks down onto the aardvark catalogs and selecting all galaxies which lie underneath. Each mask is created to accurately reproduce the HETDEX IFU pattern, see Figure~\ref{fig:ifu layout}. The pattern consists of 78 IFUs, which are comprised of 448 optical fibers subtending a $50'' \times 50''$ region on the sky \citep{Kelz2014}. The inter-IFU spacing is also $50''$ spanning a total area of $16'\times 16'$ on the sky.

The individual IFUs have a fill-factor of 1/3, which will be completely filled with three dithers of the telescope at each pointing. This means that when selecting galaxies from the aardvark catalog we assume an observation for all galaxies lying within a colored, IFU square in Figure~\ref{fig:ifu layout}. Galaxies which lie between the IFUs are missed, as well as the galaxies which lie between the pointings, as there is no overlap between one pointing and the next. This gives an overall fill-factor of 1/4.5, meaning that the entire area could be covered with 4.5 dithers of the entire survey.

\section{Recovery of Parameters}

Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

@@

There are several situations which can adversely effect the quality of our velocity dispersion measurement. The first is due to the nature of our observations. Because the HETDEX survey has a 1/4.5 fill factor (see Section~\ref{sec:Hetdex}), there will be many situations where we will not have full spectroscopic information for a cluster. Some clusters could also have intrinsic velocity distributions where the observations of only a few members could result in a LOSVD which differs by large percent. This gives rise to clusters along a fixed mass having a large range of possible LOSVDs. \cite{Saro2013} find this effect to be most pronounced for clusters with fewer than 30 members, and \cite{Ruel2014} show the effect is a net increase in scatter of about 5\% compared to clusters with greater than 30 members. \editorial{This section was adapted, pretty heavily from Ruel2014. Is that cool?}

@@

\section{Evidence of Substructure}
Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

@@

\subsection{Dressler-Schectman Test}
We leverage the large spectroscopic dataset to study the structural properties of the clusters. \cite{Pinkney1996} determine, from a comparison of five different methods that the Dressler-Shectman (DS) test \citep{Dressler1988} is the most sensitive to the presence of substructure.

@@

\section{ {\rm[\ion{O}{ii}]} Luminosity}\label{sec:oii luminosity}
HETDEX is designed to detect Lyman-$\alpha$ at redshift two. That allows us to detect [\ion{O}{ii}] ($\lambda\lambda 3727-3729\AAA$ doublet) emitters to $z\sim 0.49$. The catalog does not provide [\ion{O}{ii}] luminosities so we must assign them empirically. We use 503113 galaxies from the Sloan Digital Sky Survey (SDSS) Data Release 12 \citep{Alam2015} from $z = 0.05 - 0.2$, which are selected which have no redshift warning.

While the difference is small, we convert each SDSS galaxy's observed magnitude to a similar DES magnitude using
\begin{eqnarray}
	g_{DES} = g_{SDSS} - 0.104(g-r)_{SDSS} + 0.01 \\
	r_{DES} = r_{SDSS} - 0.102(g-r)_{SDSS} + 0.02
\end{eqnarray}
from \cite{Bechtol2015} and place each on a color-magnitude diagram (CMD) of $M_r$ and $g-r$, see Figure~\ref{fig:oii sdss}. To assign an [\ion{O}{ii}] luminosity to each galaxy in our catalog we place the catalog galaxies on the same CMD and select all SDSS galaxies in a small 2D ($M_r$, $g-r$) bin around the galaxy. We extract all of the SDSS galaxies inside that bin and create a histogram of their [\ion{O}{ii}] luminosities. Using a slice sampling technique \citep{Neal1997} we assign the catalog galaxy an [\ion{O}{ii}] luminosity based on the distribution of SDSS galaxies extracted. For catalog galaxies which are placed on the CMD near no, or very few ($1\leq n<10$) galaxies we assign it zero [\ion{O}{ii}] luminosity or the mean luminosity, respectively.

>>>>

\section{INTRODUCTION}
The Hobby Eberly Telescope Dark Energy eXperiment (HETDEX; \citealt{Hill2008}) is a trailblazing effort to observe high-redshift large scale structures using cutting edge wide-field integral field unit (IFU) spectrographs. Designed to probe the evolution of the dark energy equation of state etched onto high redshift ($z>2$) galaxies by the Baryon Acoustic Oscillations \citep{Eisenstein2005} in the first moments of the universe, the survey will observe two fields for a total of 420 \degsq\ from two fields (300 \degsq, Spring field and 120 \degsq, Fall field). Tuned to find \lya\ emitting (LAE) galaxies at $1.9<z<3.5$, HETDEX expects to find 800,000 LAEs, and more than one million [\ion{O}{ii}] emitting galaxies at $z<0.5$ masquerading as high-redshift galaxies \citep{Acquaviva2014}.

While a large portion of the $\sim10^6$ interloping [\ion{O}{ii}] galaxies will be field (not associated with a bound structure) galaxies, the large area covered by HETDEX is expected to contain as many as 100 Virgo-sized ($M_{dyn}\sim 10^{15}$ \msol) clusters at $z<0.5$ (\editorial{citation?}). The near-complete spectroscopic coverage allows an unprecedentedly detailed look at a very large number of clusters ranging from group scales to the very massive. In addition to the recovery of accurate dynamical masses, detailed investigations of the of dynamical state of the clusters is possible.

In addition to the spectroscopy, an imaging campaign is require first identify galaxy over densities. HETDEX overlaps with the Sloan Digital Sky Survey (SDSS; \citealt{Blanton2001a}), SDSS stripe 82 \citep{Annis2014}, the Dark Energy Survey (DES; \citealt{DES2005}), and the upcoming DECam/IRAC Galaxy Environment Survey (DIRGES; PI: Papovich, C. Papovich \etal in preparation).

DIRGES, in particular, will be useful as it will identify as many as 100 galaxy cluster progenitors at 0.8 $<$ z $<$ 1.5 based on similarities in galaxy \spitzer/IRAC colors. Significant gains in the understanding of these clusters are possible if we can (1) interpret their dynamical nature through spectroscopy, and (2) if we can link them to galaxy clusters at lower redshift, z $<$ 0.5, from within the same dataset to suppress potential observational biases. However, two large issues remain.

It is unclear how a blind spectroscopic survey with an IFU will effect the recovery of galaxy cluster dynamical properties. Unlike many previous large cluster surveys \citeeg{Milvang-Jensen2008, Robotham2011, Sifon2015} which use multi-object spectrographs, the Visible Integral-Field Replicable Unit Spectrograph (VIRUS; \citealt{Hill2012}) used by HETDEX samples the sky unevenly which could excluded member galaxies which would otherwise be included. Secondly, it is not straightforward to use spectroscopic redshifts predominately from emission-line galaxies to interpret the kinematic and dynamical states of the clusters.

This work plans to address these concerns in the following ways. Create and evaluate a HETDEX like selection ``function'' of galaxies over a similarly large portion of the sky, use well adopted techniques to recover the dynamical properties, such as velocity dispersion and mass, and to investigate the presence of substructure. This will better allow future work to predict the number and types of galaxy clusters which should be observed with HETDEX.

\editorial{Give outline of paper section.}

Throughout this paper, we adopt the following cosmological model ($\Omega_\Lambda = 0.77$,
$\Omega_M = 0.23$, $\sigma_8 = 0.83$ and $H_0= 72$ \kms \mpc), assume a Chabrier initial mass function (IMF; \citealt{Chabrier2003}), and use AB magnitudes \citep{Oke1974}.

%This support will allow Mr. Boada to contribute to the core science of my NSF grant (AST 1413317) for the DECam/IRAC Galaxy Environment Survey (DIRGES). He will studying both simulated and real data to understand the reliability in measuring galaxy cluster masses and dynamical states using spectroscopic information for emission-line galaxies. DIRGES will identify as many as 100 galaxy cluster progenitors at 0.8 $<$ z $<$ 1.5. Significant gains in the understanding of these clusters are possible if we can (1) interpret their dynamical nature through spectroscopy, and (2) if we can link them to clusters at lower redshift, z $<$ 0.5, from within the same dataset to suppress potential observational biases. I have plans to obtain the spectroscopy to achieve both of these supplemental goals, but the majority of the spectroscopic redshifts will be for emission-line galaxies. It is not straightforward to use these data to interpret the kinematic and dynamical states of the clusters. And, because HETDEX covers DIRGES, the data will contain both low-redshift and high-redshift clusters in the same imaging dataset, allowing the most homogeneous study of the low-z and high-z universe yet.

@@

The ``Aardvark'' mock galaxy catalogs (R. Wechsler et al., private communication) are derived from a semi-analytical model (SAM) tied to an in house n-body cosmological simulation and designed for DES. They provide a 10313 \degsq (one quarter of the sky) catalog to full DES depth and contains 1.36 billion galaxies which have a signal-to-noise of at least five in one or more DES observing bands. The catalog footprint is shown in Figure~\ref{fig:survey area}.

@@

\subsection{ {\rm[\ion{O}{ii}]} Luminosity}\label{sec:oii luminosity}
The aardvark ``truth'' catalog does not provide [\ion{O}{ii}] luminosities so we must assign them empirically. We use 503113 galaxies from the SDSS Data Release 12 \citep{Alam2015} from $z = 0.05 - 0.2$, which are selected with no redshift warning.

While the difference is small, we convert each SDSS galaxy's observed magnitude to a similar DES magnitude using
\begin{eqnarray}
	g_{DES} = g_{SDSS} - 0.104(g-r)_{SDSS} + 0.01 \nonumber \\
	r_{DES} = r_{SDSS} - 0.102(g-r)_{SDSS} + 0.02
\end{eqnarray}
from \cite{Bechtol2015} and place each on a color-magnitude diagram (CMD) of $M_r$ and $g-r$, see Figure~\ref{fig:oii sdss}. To assign an [\ion{O}{ii}] luminosity to each galaxy in our catalog we place the catalog galaxies on the same CMD and select all SDSS galaxies in a small 2D ($M_r$, $g-r$) bin around the galaxy. We extract all of the SDSS galaxies inside that bin and create a histogram of their [\ion{O}{ii}] luminosities. Using a slice sampling technique \citep{Neal1997} we assign the catalog galaxy an [\ion{O}{ii}] luminosity based on the distribution of SDSS galaxies extracted. For catalog galaxies which are placed on the CMD near no, or very few ($1\leq n<10$) galaxies we assign it zero [\ion{O}{ii}] luminosity or the mean luminosity, respectively.

\subsection{Mock Observations}\label{sec:observations}
\editorial{don't think I say what are the actual selection criteria for the spectroscopy. That's an important detail.}
Tentatively slated to start in the fall of 2015, HETDEX will perform blind spectroscopy (R $\sim$ 750 in $3500 - 5500~\AAA$) over two fields along the celestial equator. The 300 \degsq, Spring field and 120 \degsq, Fall field will have no preselected targets. Using VIRUS on the 10-m Hobby-Eberly Telescope (HET; \citealt{Ramsey1998}) the completed survey is expected to have an overall fill-factor of 1/4.5, meaning that the entire area could be covered with 4.5 dithers of the entire survey.

The spectral coverage allows for the detection of [\ion{O}{ii}] ($\lambda\lambda 3727-3729~\AAA$ doublet) emitters to $z\sim 0.5$ and Ca H ($\lambda 3968.5~\AAA$) and K ($\lambda 3933.7~\AAA$) absorption features to $z\sim 0.4$. HETDEX is expected to detect sources with continuum brighter than 22 mag in SDSS \sdssg, and emission line strengths above $3.5\times10^{-17}$ \ergscm. So we ``observe'' $z<0.4$ galaxies which meet either the emission line or the magnitude limit. Galaxies $0.4<z<0.5$ are only observed if their emission line strength is sufficient.

\begin{figure} 
	\includegraphics[width=0.49\textwidth]{f01.pdf} 
	\caption{Representative observation tiling scheme for the HETDEX $16'$ pointings. Each colored square is a single VIRUS IFU and the dashed octagons approximate the size of a single observation. See the text for more details.} \label{fig:ifu layout} 
\end{figure}

Our ``observations'' consist of placing masks down onto the aardvark ``truth'' catalogs and selecting all, $z< 0.5$ also meeting sensitivity limits, galaxies which lie underneath. Each mask is created to accurately reproduce the HETDEX IFU pattern, see Figure~\ref{fig:ifu layout}. The pattern consists of 78 IFUs, which are comprised of 448 optical fibers subtending a $50'' \times 50''$ region on the sky \citep{Kelz2014}. The inter-IFU spacing is also $50''$ spanning a total area of $16'\times 16'$ on the sky.

The individual IFUs have a fill-factor of 1/3, which will be completely filled with three dithers of the telescope at each pointing. This means that when selecting galaxies from the aardvark catalog we assume an observation for all galaxies laying within a colored, IFU square in Figure~\ref{fig:ifu layout}. Galaxies which lie between the IFUs are missed, as well as the galaxies which lie between the pointings, as there is no overlap between one pointing and the next.

\section{Cluster Membership}
In the following sections we assume that the on sky positions and redshifts of the galaxy clusters are already known. In reality, galaxy over densities will be detected from imaging surveys \citeeg{Rykoff2014} or through techniques to identify structure in the RA, DEC, z point data (\eg, DBSCAN; \citealt{Ester1996}).

While the galaxy cluster center may be well known, the constituent galaxies will not. The identification of member galaxies from photometric data \citeeg{Rozo2014} is promising, however, with spectroscopic information on all galaxies, we choose to utilize the full 3D parameter space. \editorial{Not very happy about that sentence}.

We employ the ``shifting gapped" method of \cite{Fadda1996} as outlined in \cite{Owers2011}, which combines both the positional and velocity information. After an initial cut of $cz_c \pm 10,000$ \kms, the galaxies are binned in $0.4h^{-1}$ Mpc or larger enough bins to contain at least 10 (arbitrarily chosen) galaxies. Galaxies farther than 3 Mpc from the cluster center are also rejected. Once binned, the galaxies are sorted by their line-of-sight velocity (LOSV; see Section~\ref{sec:LOSVD}). Any galaxy with a LOSV greater than 1000 \kms of a neighboring galaxy is rejected as an interloper. The procedure repeats until the number of galaxies stabilizes in the bin.

\editorial{This is taken from the other paper, and needs to be tweaked to make fit better into this framework. I think what we'll do is center a 100'' box on all of the clusters and then try to recover the galaxies. How many will we get, how many are wrong, and can we do better with a different algorithm?}

\section{Recovery of Parameters}
 In the following sections, we outline the methods we use to derive the dynamical properties of the galaxy clusters in our sample. This is not meant to be an exhaustive study of the different methods used to recover these parameters. The following is, in many cases, a subset of the available methods to derive any single parameter. The specific choice of method may improve or diminish the accuracy of the recovered parameter, but the methods chosen were to facilitate comparison with observational studies.

@@

There are several situations which can adversely effect the quality of our velocity dispersion measurement. The first is due to the nature of our observations. Because the HETDEX survey has a 1/4.5 fill factor (see Section~\ref{sec:observations}), there will be many situations where we will not have full spectroscopic information for a cluster. Some clusters could also have intrinsic velocity distributions where the observations of only a few members could result in a LOSVD which differs by large percent. This gives rise to clusters along a fixed mass having a large range of possible LOSVDs. \cite{Saro2013} find this effect to be most pronounced for clusters with fewer than 30 members, and \cite{Ruel2014} show the effect is a net increase in scatter of about 5\% compared to clusters with greater than 30 members. \editorial{This section was adapted, pretty heavily from Ruel2014. Is that cool?}

@@

\section{Evidence of Substructure}
Like the recovery of dynamical parameters discussed in the previous section, the choice of method to investigate the dynamical state of galaxy clusters is heavily debated. Perhaps the most widely used metric is the Dressler-Shectman (DS) test \citep{Dressler1988}, and is not without criticism \citeeg{White2010}. New methods such as the Caustic Method (\citealt{Yu2015}, and the references therein) offer promise to shed further light on this difficult problem.

@@

\subsection{Dressler-Schectman Test}
We leverage the large spectroscopic dataset to study the structural properties of the clusters. \cite{Pinkney1996} determine, from a comparison of five different methods that the DS test is the most sensitive to the presence of substructure.

@@
